---
title: "ONT R10 SY5Y — structural categories and UJC analysis"
author: "Fabian Jetzinger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  fl_threshold: 1
  fl_filter_levels_chr: "1,3,5,10,20"
output:
  html_document:
    toc: true
    theme: spacelab
    self_contained: true
    fig_width: 12
    fig_height: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 12,
  fig.height = 10
)

library(grid)
library(tidyverse)
library(cowplot)
library(patchwork)
library(scales)
library(RColorConesa)
library(ComplexUpset)

r_source_dir <- if (dir.exists("R")) "R" else "src/plotting/R"

source(file.path(r_source_dir, "02_themes.R"))
source(file.path(r_source_dir, "03_plot_functions.R"))
source(file.path(r_source_dir, "01_config_ont_r10.R"))
source(file.path(r_source_dir, "07_ont_r10_sy5y.R"))

parse_fl_levels <- function(s) {
  v <- as.numeric(trimws(strsplit(s, ",", fixed = TRUE)[[1L]]))
  v[!is.na(v)]
}

fl_levels <- parse_fl_levels(params$fl_filter_levels_chr)
if (!length(fl_levels)) {
  fl_levels <- ont_r10_sy5y_fl_filter_levels
}
thr_ujc <- as.numeric(params$fl_threshold)

out_dir <- sy5y_plot_creation_dir()
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

```{r check-paths, results="asis"}
check_paths_ont_r10()
```

```{r load-results-fl1}
all_results <- sy5y_build_all_tools_results(thr_ujc, include_expression = FALSE)
```

```{r ylim-diagnostics, echo=TRUE, results="asis"}
sy5y_print_ylim_diagnostics(all_results)
```

## Structural categories (FL filter)

Same count-based view as mouse `all_plots.Rmd` (no stacked “% of isoforms” panel). One **2×2** PDF per FL threshold.

```{r structural-loop, results="hide"}
for (fl in fl_levels) {
  iso_p <- sy5y_classification_count_panel(
    all_results$IsoQuant$raw, sy5y_sample_labels(), fl
  )
  fl_p <- sy5y_classification_count_panel(
    all_results$FLAIR$raw, sy5y_sample_labels(), fl
  )
  bm_p <- sy5y_classification_count_panel(
    all_results$Bambu$raw, sy5y_sample_labels(), fl
  )
  title <- sprintf(
    "ONT R10 SY5Y — transcripts per structural category (filter: FL >= %s)",
    fl
  )
  p <- assemble_ont_r10_tool_grid(iso_p, fl_p, bm_p, title = title)
  ggsave(
    file.path(out_dir, sprintf("sy5y_structural_category_fl%s.pdf", fl)),
    plot = p,
    device = grDevices::cairo_pdf,
    width = 14,
    height = 11,
    units = "in"
  )
}
```

```{r show-structural, echo=FALSE}
fl_show <- if (thr_ujc %in% fl_levels) thr_ujc else fl_levels[[1L]]
iso_show <- sy5y_classification_count_panel(
  all_results$IsoQuant$raw, sy5y_sample_labels(), fl_show
)
fl_show_p <- sy5y_classification_count_panel(
  all_results$FLAIR$raw, sy5y_sample_labels(), fl_show
)
bm_show <- sy5y_classification_count_panel(
  all_results$Bambu$raw, sy5y_sample_labels(), fl_show
)
assemble_ont_r10_tool_grid(
  iso_show,
  fl_show_p,
  bm_show,
  title = sprintf(
    "ONT R10 SY5Y — transcripts per structural category (filter: FL >= %s)",
    fl_show
  )
)
```

---

## Expression (FL >= 1)

Transcript read support per structural category (with unassigned reads), mirroring mouse `assemble_expr_bar_plots` for ONT only.

```{r expression-build}
for (tool in SY5Y_TOOL_ORDER) {
  all_results[[tool]]$expression <- sy5y_expression_panel(
    all_results[[tool]]$raw,
    fl_threshold = 1L
  )
}
expr_grid <- assemble_ont_r10_expression_grid(
  all_results,
  title = "ONT R10 SY5Y — transcript expression per structural category"
)
```

```{r expression-save, results="hide"}
ggsave(
  file.path(out_dir, "sy5y_expression_fl1.pdf"),
  plot = expr_grid,
  device = grDevices::cairo_pdf,
  width = 14,
  height = 11,
  units = "in"
)
```

```{r show-expression, echo=FALSE}
expr_grid
```

---

## UJC analysis — multi-tool 2×2 grids (FL >= `r thr_ujc`)

Panels match mouse `all_plots.Rmd` (non-percent only). Per-tool PDFs and HTML overviews are kept for inspection.

```{r ujc-titles}
ujc_titles <- list(
  upset = "Unique Junction Chain overlap; ONT R10 SY5Y",
  comb = "Unique Junction Chain occurrence across samples; ONT R10 SY5Y",
  comb_bar = "Unique Junction Chain occurrence across samples; ONT R10 SY5Y",
  comb_fl_bar = "Expression of UJCs across samples; ONT R10 SY5Y",
  ujc_curve = "Transcript occurrence across number of samples; ONT R10 SY5Y",
  ujc_stack = "UJC occurrence across combinations of samples; ONT R10 SY5Y",
  ujc_fl_stack = "Expression of UJCs across combinations of samples; ONT R10 SY5Y"
)
ujc_ylabs <- list(
  comb = "Number of UJCs",
  comb_bar = "# of UJCs",
  comb_fl_bar = "# of reads of UJCs",
  ujc_stack = "Mean # of UJCs",
  ujc_fl_stack = "Mean # of reads of UJCs"
)
```

### UpSet

```{r ujc-upset-save, results="hide"}
p_upset <- assemble_ont_r10_upset_grid(all_results, title = ujc_titles$upset)
ggsave(
  file.path(out_dir, sprintf("sy5y_upset_fl%s.pdf", thr_ujc)),
  plot = p_upset,
  device = grDevices::cairo_pdf,
  width = 16,
  height = 12,
  units = "in"
)
```

### UJC occurrence across samples — transcript counts

Mouse: `brain_comb_bar.pdf` (`# of UJCs` on the y-axis).

```{r ujc-occurrence-transcripts-save, results="hide"}
sy5y_ggsave_ujc_multitool_panel(
  all_results,
  SY5Y_UJC_OCCURRENCE_TRANSCRIPT,
  title = ujc_titles$comb_bar,
  out_dir = out_dir,
  fl_threshold = thr_ujc,
  ylab = ujc_ylabs$comb_bar
)
```

```{r show-ujc-comb-bar, echo=FALSE}
assemble_ont_r10_ujc_ggplot_grid(
  all_results,
  SY5Y_UJC_OCCURRENCE_TRANSCRIPT,
  title = ujc_titles$comb_bar,
  ylab = ujc_ylabs$comb_bar
)
```

### UJC occurrence across samples — read support (FL)

Mouse: `brain_comb_fl_bar.pdf` (`# of reads of UJCs` on the y-axis). Same x-axis (number of replicate samples present), summed FL read support per structural category.

```{r ujc-occurrence-reads-save, results="hide"}
sy5y_ggsave_ujc_multitool_panel(
  all_results,
  SY5Y_UJC_OCCURRENCE_READS,
  title = ujc_titles$comb_fl_bar,
  out_dir = out_dir,
  fl_threshold = thr_ujc,
  ylab = ujc_ylabs$comb_fl_bar,
  ylims = ont_r10_sy5y_ylims$comb_fl_bar
)
```

```{r show-ujc-comb-fl-bar, echo=FALSE}
assemble_ont_r10_ujc_ggplot_grid(
  all_results,
  SY5Y_UJC_OCCURRENCE_READS,
  title = ujc_titles$comb_fl_bar,
  ylab = ujc_ylabs$comb_fl_bar,
  ylims = ont_r10_sy5y_ylims$comb_fl_bar
)
```

### Other UJC panels

```{r ujc-other-save, results="hide"}
ujc_other_panels <- setdiff(
  SY5Y_UJC_PANEL_NAMES,
  c("upset", SY5Y_UJC_OCCURRENCE_TRANSCRIPT, SY5Y_UJC_OCCURRENCE_READS)
)
for (pn in ujc_other_panels) {
  sy5y_ggsave_ujc_multitool_panel(
    all_results,
    pn,
    title = ujc_titles[[pn]],
    out_dir = out_dir,
    fl_threshold = thr_ujc,
    ylab = ujc_ylabs[[pn]]
  )
}
```

```{r ujc-per-tool-save, results="hide"}
for (tool in SY5Y_TOOL_ORDER) {
  slug <- tolower(tool)
  sy5y_save_ujc_pdfs(all_results[[tool]]$ujc, slug, out_dir)
  ggsave(
    file.path(out_dir, sprintf("sy5y_ujc_overview_%s_fl%s.pdf", slug, thr_ujc)),
    plot = sy5y_ujc_overview_patchwork(
      all_results[[tool]]$ujc,
      paste0(tool, " — UJC overview")
    ),
    device = grDevices::cairo_pdf,
    width = 16,
    height = 22,
    units = "in"
  )
}
```

```{r show-ujc-upset, echo=FALSE, fig.height=12, fig.width=16}
assemble_ont_r10_upset_grid(all_results, title = ujc_titles$upset)
```

---

## Join&Call vs Call&Join — multi-tool redundancy (FL >= `r thr_ujc`)

Four **2×2** figures (one per panel type), matching mouse `assemble_compare_plots`.

```{r compare-multitool-save, results="hide"}
cmp_titles <- list(
  count_transcript = "Redundancy of transcripts (# of transcripts); SY5Y",
  total_isoforms_transcript = "Redundancy of transcripts (# of isoforms); SY5Y",
  count_ujc = "Redundancy of UJCs (# of UJCs); SY5Y",
  total_isoforms_ujc = "Redundancy of UJCs (# of isoforms); SY5Y"
)
for (pn in SY5Y_COMPARE_PANEL_NAMES) {
  p <- assemble_ont_r10_compare_grid(all_results, pn, title = cmp_titles[[pn]])
  ggsave(
    file.path(out_dir, sprintf("sy5y_compare_%s_fl%s.pdf", pn, thr_ujc)),
    plot = p,
    device = grDevices::cairo_pdf,
    width = 14,
    height = 11,
    units = "in"
  )
}
```

```{r compare-per-tool-save, results="hide"}
for (tool in SY5Y_TOOL_ORDER) {
  slug <- tolower(tool)
  sy5y_save_compare_pdfs(all_results[[tool]]$compare, slug, out_dir)
  ggsave(
    file.path(out_dir, sprintf("sy5y_compare_overview_%s_fl%s.pdf", slug, thr_ujc)),
    plot = sy5y_compare_patchwork(all_results[[tool]]$compare) +
      patchwork::plot_annotation(
        title = paste0(tool, " — Join&Call vs Call&Join"),
        theme = ggplot2::theme(
          plot.title = ggplot2::element_text(face = "bold", hjust = 0.5, size = 14)
        )
      ),
    device = grDevices::cairo_pdf,
    width = 14,
    height = 12,
    units = "in"
  )
}
```

```{r show-compare, echo=FALSE, fig.width=14, fig.height=11}
assemble_ont_r10_compare_grid(
  all_results,
  "count_transcript",
  title = cmp_titles$count_transcript
)
```

---

## Output summary

Multi-tool PDFs under `plot_creation_ont_r10/`:

- `sy5y_structural_category_fl{1,3,5,10,20}.pdf`
- `sy5y_expression_fl1.pdf`
- `sy5y_upset_fl1.pdf`
- **UJC occurrence across samples:** `sy5y_comb_bar_fl1.pdf` (# UJCs) and `sy5y_comb_fl_bar_fl1.pdf` (# reads of UJCs; mouse `brain_comb_fl_bar.pdf`)
- `sy5y_comb_fl1.pdf` (boxplot), `sy5y_ujc_curve_fl1.pdf`, `sy5y_ujc_stack_fl1.pdf`, `sy5y_ujc_fl_stack_fl1.pdf`
- `sy5y_compare_<panel>_fl1.pdf` (four compare panels)
- Per-tool: `sy5y_<tool>_ujc_*.pdf`, `sy5y_ujc_overview_<tool>_fl1.pdf`, `sy5y_<tool>_compare_*.pdf`

TPM filter figures: knit [`filter_level_ont_r10_sy5y_plots.Rmd`](filter_level_ont_r10_sy5y_plots.Rmd).

Read-level figures: [`read_level_ont_r10_sy5y_plots.Rmd`](read_level_ont_r10_sy5y_plots.Rmd).
