---
title: "ONT R10 SY5Y — read-level plots (SQANTI-reads, read counts, BAM lengths)"
author: "Fabian Jetzinger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  sample_ids_chr: "SRR31732191,SRR31732198,SRR31732209,SRR31732210"
  sample_labels_chr: "191,198,209,210"
  # SQANTI "Unaligned" slice: "primary_mapped" uses ont_prim_aln (recommended for SY5Y
  # because SQANTI is run on BAM->GTF input; see diagnostics chunk). "fastq" matches
  # mouse read_figure_plots.Rmd (ont_fastq - assigned).
  sqanti_unaligned_denominator: "primary_mapped"
output:
  html_document:
    toc: true
    theme: spacelab
    self_contained: true
    fig_width: 12
    fig_height: 8
---

Read-level figures aligned with mouse `read_figure_plots.Rmd`: SQANTI-reads structural
categories, stacked FASTQ vs primary-mapped counts, and BAM read-length violins
(99.5% quantile trim). **ONT only** (four SY5Y replicates).

### Mouse (NIH) vs this R10 pipeline (important)

| Step | Mouse (repo / `read_figure_plots.Rmd`) | R10 SY5Y (this repo) |
|------|----------------------------------------|----------------------|
| ONT SQANTI-reads merge | `sqanti_reads.py` + `design_ont.csv`; per-sample `*_reads_classification.txt` under NIH paths (`02_sqanti_reads_ONT.sbatch`) | `6_sqanti_reads.sh`: **primary BAM → `spliced_bam2gff` → GTF → `sqanti3_qc.py`** → `*_reads_classification.txt`; then `7_sqanti_reads_merge.sh` runs `sqanti_reads.py` on merged design |
| Plot reads ONT file names | `ont_dir`: `B31_classification.txt` … (no `_reads_` in Rmd; `design_ont.csv` uses `*_reads_classification.txt` — **local mirror may rename**) | `*_reads_classification.txt` under `sqanti_reads/<SRR>/` |
| Read totals / lengths | `count_reads.sbatch` + `get_read_lengths.sbatch` on NIH metadata TSVs | `9_count_reads_joint.sh` + `10_get_read_lengths_bam.sh` + same column layout as mouse TSV |
| Unaligned in SQANTI stack (R) | `ont_fastq - sum(classified reads)` in `build_category_counts()` | **Default here: `ont_prim_aln - assigned`** (`params$sqanti_unaligned_denominator: primary_mapped`) because classification rows come from **BAM-derived GTF**, not necessarily one row per FASTQ read. Set to `fastq` to reproduce the **literal mouse R formula** on this data (often inflates Unaligned). |

The HTML **diagnostics** chunk prints: classification file row count, rows that become `NA` after category recode (excluded from stacked sums), `ont_fastq` / `ont_prim_aln`, **`n_primary_mapped_minus_rows`** (primary BAM reads not represented as a classification row), and **`pct_sqanti_rows_of_primary_mapped`**.

**Expected layout under the repository root** (e.g. after `rsync` from the cluster):

- `data/ont_r10/read_qc/read_numbers_joint.tsv` — from `src/preprocessing/ont_r10/9_count_reads_joint.sh`
- `data/ont_r10/read_qc/lengths_seq_only/ont/<SRR>.bam.readlen.txt` — from `10_get_read_lengths_bam.sh`
- `data/ont_r10/sqanti_reads/<SRR>/<SRR>_reads_classification.txt` — SQANTI-reads outputs

Override paths with env var `JOIN_AND_CALL_REPO` if the project is not found relative to this Rmd.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 12,
  fig.height = 8
)

required_cran <- c(
  "tidyverse",
  "ggplot2",
  "scales",
  "readr",
  "patchwork",
  "cowplot",
  "RColorConesa",
  "stringr",
  "forcats"
)
missing_cran <- required_cran[!required_cran %in% utils::installed.packages()[, "Package"]]
if (length(missing_cran) > 0) {
  message("Installing missing CRAN packages: ", paste(missing_cran, collapse = ", "))
  utils::install.packages(missing_cran, repos = "https://cloud.r-project.org")
}
```

```{r libraries}
library(tidyverse)
library(scales)
library(readr)
library(patchwork)
library(cowplot)
library(RColorConesa)
library(stringr)
library(forcats)

r_source_dir <- if (dir.exists("R")) "R" else "src/plotting/R"
source(file.path(r_source_dir, "01_config_ont_r10.R"))
source(file.path(r_source_dir, "08_read_level_sy5y.R"))

parse_csv_words <- function(s) {
  v <- strsplit(s, ",", fixed = TRUE)[[1L]]
  v <- trimws(v)
  v[nzchar(v)]
}

sample_ids <- parse_csv_words(params$sample_ids_chr)
sample_labels <- parse_csv_words(params$sample_labels_chr)
if (length(sample_ids) != length(sample_labels)) {
  stop("params$sample_ids_chr and params$sample_labels_chr must have the same number of entries.")
}

out_dir <- file.path(ont_r10_repo_root, "plot_creation_ont_r10")
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

```{r check-paths, results="asis"}
check_paths_ont_r10_read_level(sample_ids = sample_ids)
```

```{r load-read-numbers}
read_number_file <- file.path(ont_r10_read_qc_dir, "read_numbers_joint.tsv")
if (!file.exists(read_number_file)) {
  stop("Missing ", read_number_file)
}
read_numbers <- read.table(read_number_file, header = TRUE, sep = "\t", check.names = FALSE)
```

```{r sqanti-diagnostics, echo=FALSE, results="asis"}
diag_tbl <- read_level_sy5y_sqanti_diagnostics(
  read_numbers,
  sample_ids = sample_ids,
  sample_labels = sample_labels,
  sqanti_reads_root = ont_r10_sqanti_reads_root
)
knitr::kable(
  diag_tbl,
  caption = "SQANTI-reads classification vs read_numbers (per replicate)"
)
```

```{r sqanti-category-counts}
ud <- tolower(trimws(as.character(params$sqanti_unaligned_denominator)))
if (!ud %in% c("primary_mapped", "fastq")) {
  stop("params$sqanti_unaligned_denominator must be 'primary_mapped' or 'fastq'")
}
unaligned_denom <- if (ud == "fastq") "ont_fastq" else "ont_prim_aln"

category_counts_sy5y <- build_sy5y_sqanti_category_counts(
  read_numbers,
  sample_ids = sample_ids,
  sample_labels = sample_labels,
  sqanti_reads_root = ont_r10_sqanti_reads_root,
  unaligned_denominator = unaligned_denom
)

sq_title <- sprintf(
  "ONT R10 SY5Y — SQANTI-reads (Unaligned denominator: %s)",
  if (unaligned_denom == "ont_fastq") "ont_fastq (mouse Rmd parity)" else "ont_prim_aln (BAM-aligned total)"
)
sq_reads_plot <- plot_read_level_sqanti_faceted(category_counts_sy5y, title = sq_title)
sq_reads_plot
```

```{r save-sqanti-pdf, results="hide"}
ggsave(
  file.path(out_dir, "sy5y_sq_reads_plot.pdf"),
  plot = sq_reads_plot,
  device = grDevices::cairo_pdf,
  width = 14,
  height = 10,
  units = "in"
)
```

```{r readnum-plot}
readnum_tidy <- build_sy5y_readnum_tidy(read_numbers, sample_ids, sample_labels)
readnum_plot <- plot_sy5y_readnum_faceted(readnum_tidy, title = "ONT R10 SY5Y — FASTQ vs primary-mapped (ONT)")
readnum_plot
```

```{r save-readnum-pdf, results="hide"}
ggsave(
  file.path(out_dir, "sy5y_readnum_plot.pdf"),
  plot = readnum_plot,
  device = grDevices::cairo_pdf,
  width = 14,
  height = 10,
  units = "in"
)
```

```{r lengths}
length_ont_dir <- file.path(ont_r10_read_qc_dir, "lengths_seq_only", "ont")
all_lengths_df <- build_sy5y_lengths_df(length_ont_dir, sample_ids, sample_labels)
lengths_violin <- plot_sy5y_lengths_violin(all_lengths_df, title = "ONT R10 SY5Y — read length (BAM SEQ)")
lengths_violin
```

```{r save-lengths-pdf, results="hide"}
ggsave(
  file.path(out_dir, "sy5y_lengths_violin.pdf"),
  plot = lengths_violin,
  device = grDevices::cairo_pdf,
  width = 14,
  height = 10,
  units = "in"
)
```

```{r combined-patchwork}
sy5y_read_level_combined <- patchwork::wrap_plots(
  L = sq_reads_plot +
    theme(legend.position = "right", plot.title = element_blank(), plot.margin = margin(r = 12)),
  R = lengths_violin +
    theme(plot.title = element_blank(), plot.margin = margin(l = 12)),
  design = "LR"
)
sy5y_read_level_combined
```

```{r save-combined-pdf, results="hide"}
ggsave(
  file.path(out_dir, "sy5y_read_level_combined.pdf"),
  plot = sy5y_read_level_combined,
  device = grDevices::cairo_pdf,
  width = 16,
  height = 10,
  units = "in"
)
```

PDFs written under `plot_creation_ont_r10/`:

- `sy5y_sq_reads_plot.pdf`
- `sy5y_readnum_plot.pdf`
- `sy5y_lengths_violin.pdf`
- `sy5y_read_level_combined.pdf`

Render from the repository:

```r
rmarkdown::render("src/plotting/read_level_ont_r10_sy5y_plots.Rmd")
```

Or Knit this file in RStudio (working directory defaults to the folder containing the Rmd).
