---
title: "SY5Y; read-level plots (ONT R10)"
author: "Fabian Jetzinger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  sample_ids_chr: "SRR31732191,SRR31732198,SRR31732209,SRR31732210"
  sample_labels_chr: "191,198,209,210"
output:
  html_document:
    toc: true
    theme: spacelab
    self_contained: true
    fig_width: 12
    fig_height: 8
---

Read-level figures aligned with mouse `read_figure_plots.Rmd`: SQANTI-reads structural
categories and primary-alignment read-length violins. **ONT only** (four SY5Y replicates).
Plot styling is shared via `08_read_level_plots.R` (same as mouse).

### Mouse (NIH) vs this R10 pipeline (important)

| Step | Mouse (repo / `read_figure_plots.Rmd`) | R10 SY5Y (this repo) |
|------|----------------------------------------|----------------------|
| ONT SQANTI-reads merge | `sqanti_reads.py` + `design_ont.csv`; per-sample `*_reads_classification.txt` under NIH paths (`02_sqanti_reads_ONT.sbatch`) | `6_sqanti_reads.sh`: **primary BAM → `spliced_bam2gff` → GTF → `sqanti3_qc.py`** → `*_reads_classification.txt`; then `7_sqanti_reads_merge.sh` runs `sqanti_reads.py` on merged design |
| Plot reads ONT file names | `ont_dir`: `B31_classification.txt` … (no `_reads_` in Rmd; `design_ont.csv` uses `*_reads_classification.txt` — **local mirror may rename**) | `*_reads_classification.txt` under `sqanti_reads/<SRR>/` |
| Read totals / lengths | `count_reads.sbatch` + `get_read_lengths.sbatch` on NIH metadata TSVs | `9_count_reads_joint.sh` + `10_get_read_lengths_bam.sh` + same column layout as mouse TSV |
| Extra read buckets (R) | Single "Unaligned" in mouse | **Unaligned** = `ont_fastq - ont_prim_aln`; **Unstranded** = `ont_prim_aln - SQANTI-classified` (mapped but no classification row). |
| Expression "Unassigned" total | `ont_fastq` | `ont_fastq` (unchanged; see `ont_r10_sy5y_plots.Rmd`) |

The HTML **diagnostics** chunk prints: classification file row count, rows that become `NA` after category recode (excluded from stacked sums), `ont_fastq` / `ont_prim_aln`, **`n_primary_mapped_minus_rows`** (primary BAM reads not represented as a classification row), and **`pct_sqanti_rows_of_primary_mapped`**.

**Expected layout under the repository root** (e.g. after `rsync` from the cluster):

- `data/ont_r10/read_qc/read_numbers_joint.tsv` — from `src/preprocessing/ont_r10/9_count_reads_joint.sh`
- `data/ont_r10/read_qc/lengths_seq_only/ont/<SRR>.bam.readlen.txt` — from `10_get_read_lengths_bam.sh`
- `data/ont_r10/sqanti_reads/<SRR>/<SRR>_reads_classification.txt` — SQANTI-reads outputs

Override paths with env var `JOIN_AND_CALL_REPO` if the project is not found relative to this Rmd.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)

required_cran <- c(
  "tidyverse",
  "ggplot2",
  "scales",
  "readr",
  "patchwork",
  "cowplot",
  "RColorConesa",
  "stringr",
  "forcats"
)
missing_cran <- required_cran[!required_cran %in% utils::installed.packages()[, "Package"]]
if (length(missing_cran) > 0) {
  message("Installing missing CRAN packages: ", paste(missing_cran, collapse = ", "))
  utils::install.packages(missing_cran, repos = "https://cloud.r-project.org")
}
```

```{r libraries}
library(tidyverse)
library(scales)
library(readr)
library(patchwork)
library(cowplot)
library(RColorConesa)
library(stringr)
library(forcats)

r_source_dir <- if (dir.exists("R")) "R" else "src/plotting/R"
source(file.path(r_source_dir, "00_figure_config.R"))
source(file.path(r_source_dir, "02_themes.R"))
source(file.path(r_source_dir, "08_read_level_plots.R"))
source(file.path(r_source_dir, "01_config_ont_r10.R"))
source(file.path(r_source_dir, "08_read_level_sy5y.R"))

parse_csv_words <- function(s) {
  v <- strsplit(s, ",", fixed = TRUE)[[1L]]
  v <- trimws(v)
  v[nzchar(v)]
}

sample_ids <- parse_csv_words(params$sample_ids_chr)
sample_labels <- parse_csv_words(params$sample_labels_chr)
if (length(sample_ids) != length(sample_labels)) {
  stop("params$sample_ids_chr and params$sample_labels_chr must have the same number of entries.")
}

out_dir <- plot_creation_ont_r10_dir()
```

```{r check-paths, results="asis"}
check_paths_ont_r10_read_level(sample_ids = sample_ids)
```

```{r load-read-numbers}
read_number_file <- file.path(ont_r10_read_qc_dir, "read_numbers_joint.tsv")
if (!file.exists(read_number_file)) {
  stop("Missing ", read_number_file)
}
read_numbers <- read.table(read_number_file, header = TRUE, sep = "\t", check.names = FALSE)
```

```{r sqanti-diagnostics, echo=FALSE, results="asis"}
diag_tbl <- read_level_sy5y_sqanti_diagnostics(
  read_numbers,
  sample_ids = sample_ids,
  sample_labels = sample_labels,
  sqanti_reads_root = ont_r10_sqanti_reads_root
)
knitr::kable(
  diag_tbl,
  caption = "SQANTI-reads classification vs read_numbers (per replicate)"
)
```

```{r sqanti-category-counts}
category_counts_sy5y <- build_sy5y_sqanti_category_counts(
  read_numbers,
  sample_ids = sample_ids,
  sample_labels = sample_labels,
  sqanti_reads_root = ont_r10_sqanti_reads_root
)

sq_reads_plot <- plot_read_level_sqanti_faceted(
  category_counts_sy5y,
  title = ont_r10_sy5y_condition_label
)
sq_reads_plot
paper_ggsave(
  file.path(out_dir, "sy5y_sq_reads_plot.pdf"),
  plot = sq_reads_plot,
  size = "sy5y_read_level_panel"
)
```

```{r read-lengths}
length_ont_dir <- file.path(ont_r10_read_qc_dir, "lengths_seq_only", "ont")
lengths_df <- build_sy5y_lengths_df(length_ont_dir, sample_ids, sample_labels)
lengths_violin <- plot_sy5y_lengths_violin(
  lengths_df,
  title = ont_r10_sy5y_condition_label
)
lengths_violin
paper_ggsave(
  file.path(out_dir, "sy5y_lengths_violin.pdf"),
  plot = lengths_violin,
  size = "sy5y_read_level_panel"
)
```

```{r combined-patchwork}
sy5y_read_level_combined <- assemble_sy5y_read_level_combined(
  sq_reads_plot,
  lengths_violin
)
sy5y_read_level_combined
```

```{r save-combined-pdf, results="hide"}
paper_ggsave(
  file.path(out_dir, "sy5y_read_level_combined.pdf"),
  plot = sy5y_read_level_combined,
  size = "sy5y_read_level_combo"
)
```

```{r source_data_capture, echo=FALSE, message=FALSE, warning=FALSE}
# ---- Source Data capture: SY5Y read-level figure (SF22, former SRF4) ---------
# Two sub-sheets, mirroring mouse F3/SF1:
#   a = stacked structural-category read counts (already aggregated).
#   b = read-length distribution -> summary stats (n + five-number summary, nt)
#       on the same 99.5th-percentile-trimmed reads the violin shows (raw per-
#       read lengths are millions of rows).
.sd_sy5y_length_summary <- function(lengths_df, trim_quantile = 0.995) {
  lengths_df %>%
    dplyr::group_by(.data$sample, .data$technology) %>%
    dplyr::filter(
      .data$length <= stats::quantile(.data$length, trim_quantile, na.rm = TRUE)
    ) %>%
    dplyr::summarise(
      n_reads   = dplyr::n(),
      min_nt    = min(.data$length),
      q25_nt    = stats::quantile(.data$length, 0.25),
      median_nt = stats::median(.data$length),
      mean_nt   = mean(.data$length),
      q75_nt    = stats::quantile(.data$length, 0.75),
      max_nt    = max(.data$length),
      .groups   = "drop"
    ) %>%
    as.data.frame()
}

add_source_sheet("Supplementary Fig 22a",
  "Supplementary Fig. 22a: number of reads per SQANTI3 structural category, per replicate (SH-SY5Y, ONT R10); includes Unaligned and Unstranded read buckets.",
  as.data.frame(category_counts_sy5y))
add_source_sheet("Supplementary Fig 22b",
  "Supplementary Fig. 22b: read-length distribution per replicate (SH-SY5Y, ONT R10); summary statistics (nt) of the 99.5th-percentile-trimmed reads shown as violin + box plots.",
  .sd_sy5y_length_summary(lengths_df))
```

PDF written under `plot_creation_ont_r10/`:

- `sy5y_sq_reads_plot.pdf`, `sy5y_lengths_violin.pdf`
- `sy5y_read_level_combined.pdf` (SQANTI-reads stack + read-length violins)

Render from the repository:

```r
rmarkdown::render("src/plotting/read_level_ont_r10_sy5y_plots.Rmd")
```

Or Knit this file in RStudio (working directory defaults to the folder containing the Rmd).
